\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-14 coqimpl.tex
Fix typo.

\begin{figure}[!htb]
  \begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
(* ... *)
  Family subst_internal. 
  Final Field tm_var : ...
:= fun s x t => 
    if (eqb x s) 
    then t else (tm_var s).
  Final Field tm_abs : ...
:= fun s (b : tm)   
   (recb : id -> tm -> tm) 
   x t => 
    if (eqb x s) 
    then (tm_abs s body)
    else (tm_abs s (recb x t)).
(* ... *)
  EndFamily.
  FRecursor subst 
    about tm 
    motive (fun _ => id -> tm -> tm)
    using subst_internal
    by _rec.
(*
Field test := (subst (tm_var 0) 0 (tm_var 0)).
*)
(* ... *)
EndFamily.
Family STLC_bool extends STLC.
(* ... *)
Extend FInductive tm : Set :=
  | tm_true : tm ...

@@

Module Type STLC_409.
  Parameter (tm : Set).
  Parameter (tm_var : id -> tm).
  Parameter (tm_abs : id -> tm -> tm).
  (* ... and all the fields defined before
  and the abstraction for inductive type ...*)
End STLC_409.
(* Collect eliminator to Set for tm *)
Module tm_rec_425 (STLC: STLC_409).
  Definition __recursor_type_tm_rec :=
    forall P : STLC.tm -> Set,
    (forall n : id, P (STLC.tm_var n)) ->
    (forall (n : id) (i : STLC.tm),
    P i -> P (STLC.tm_abs n i)) ->
    ... -> forall i : STLC.tm, P i.
End tm_rec_425.
(* Compiled Field STLC.subst_internal.tm_var *)
Module tm_var_411 (STLC: STLC_409)
  (subst_internal: EmptySig).
  Definition tm_var :
    forall (s : id) (x : id) (t : STLC.tm), STLC.tm :=
    fun s x t => if eqb x s then t else STLC.var s.
End tm_var_411. (* ... and more *)
(* Aggregate into one subst_internal*)
Module subst_internal_410 (STLC: STLC_409). 
  Include tm_var_411 STLC. (* And other cases ... *)
End subst_internal_410.
Module v_33_34 (STLC: STLC_409).
(* Intermediate Module solely for type checking *)
  Include tm_rec_425 STLC.
  Parameter (recursor_for_type_checking : __recursor_type_tm_rec).
  Definition term_for_type_checking :=
    recursor_for_type_checking (fun _  => id -> STLC.tm -> STLC.tm)
      subst_internal_410.tm_var
      subst_internal_410.tm_abs ... .
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Exempler STLC Code, especially about Inductive Type}\label{fig:plugin-example2}
\end{figure}\YZ{Change this one into STLC example, to see the layout}\EDJreply{Changed. I think it only make things worse.}

@@

However, we later still need the concrete eliminator on "tm" for doing exhaustiveness checking (as in \ref{chg:extensible-inductive-type}), thus we will store the corresponding info somewhere.  Here we show one exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module "tm_rec_425". This concrete eliminator will be used by our "FRecursor". For example, in \cref{fig:plugin-example2}, "FRecursor" command will feed Coq a module "v_33_34" for doing type-checking with specified handlers (e.g. "neg_handler") during compilation---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq. Note that, at this point of the example, "FRecursor" will only do type check. The compilation of "FRecursor" into a component of the compiled module only happens when closing the whole family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id -> tm -> tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla) inductive type in that compiled module. Thus a "FRecursor" field is unlike other fields, even though it is exposed as a mundane function for the following fields in the programming interface.

@@

Importantly, the recursor field (e.g. "subst") is not actually inheritable, because it is not compiled like any other fields---recall how other fields are compiled into reusable pieces of (parametrized) modules, a recursor field is not compiled like that. For all the children family, we have to reconstruct and override the "FRecursor" field with a ``new'' recursor using (maybe the same) handler family. This is affordable since each aggregated recursor handler doesn't need
to be rechecked---look at "STLC_bool.neg_handler", there are only  new
handlers for "tm_true" and the handlers (mundane fields) from the parent
"B.subst_internal.tm_var" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "STLC_409"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "tm_var_411"---doesn't require another type-check and is re-used during the type-checking and compilation of "STLC_bool.subst". For example, for the new "subst", when compiling family "STLC_bool" into a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id -> tm -> tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla but extended and modified) inductive type in that compiled module. Note that, even though "tm_rec" and "tm" has the same name as before, they are in the newly-compiled module so they refer to totally different things.

@@

\begin{minted}[escapeinside=@@]{Coq}
b_prec : forall R, option R -> option R -> b -> option R.@\YZ{Not clear what it takes to define/prove partial recursors. The reader needs to be told if they require breaking abstraction like recursors do.}\EDJreply{What do you mean? recursor doesn't break abstraction but break extensibility. Below mentioned partial recursor doesn't break extensibility}@
\end{minted}
with appropriate computational axioms. This partial recursor differs from original recursor only by decorating an "option" with the return type thus possible to be automatically generated (and the case is similar for those computational axiom). The partiality doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec") because "b_prec" can just return "None" when bump into extended constructor in the future.

The first thing to check is that, if partial recursor is theoretically enough for injectiveness of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{self_B.tt ≠
self_B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : self_B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : self_B.tt = self_B.ff → False.
(* Because self_B.tt = self_B.ff → reflect self_B.tt = reflect self_B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

>>>>

\begin{figure}[!htb]
  \begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
(* ... *)
  Family subst_internal. 
  Final Field tm_var : ...
:= fun s x t => 
    if (eqb x s) 
    then t else (tm_var s).
  Final Field tm_abs : ...
:= fun s (b : tm)   
   (recb : id → tm → tm) 
   x t => 
    if (eqb x s) 
    then (tm_abs s body)
    else (tm_abs s (recb x t)).
(* ... *)
  EndFamily.
  FRecursor subst 
    about tm 
    motive (fun _ => id → tm → tm)
    using subst_internal
    by _rec.
(*
Field test := (subst (tm_var 0) 0 (tm_var 0)).
*)
(* ... *)
EndFamily.
Family STLC_bool extends STLC.
(* ... *)
Extend FInductive tm : Set :=
  | tm_true : tm ...

@@

Module Type STLC_409.
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* ... and all the fields defined before
  and the abstraction for inductive type ...*)
End STLC_409.
(* Collect eliminator to Set for tm *)
Module tm_rec_425 (STLC: STLC_409).
  Definition __recursor_type_tm_rec :=
    forall P : STLC.tm -> Set,
    (forall n : id, P (STLC.tm_var n)) ->
    (forall (n : id) (i : STLC.tm),
    P i -> P (STLC.tm_abs n i)) ->
    ... -> forall i : STLC.tm, P i.
End tm_rec_425.
(* Compiled Field STLC.subst_internal.tm_var *)
Module tm_var_411 (STLC: STLC_409)
  (subst_internal: EmptySig).
  Definition tm_var :
    forall (s : id) (x : id) (t : STLC.tm), STLC.tm :=
    fun s x t ⇒ if eqb x s then t else STLC.var s.
End tm_var_411. (* ... and more *)
(* Aggregate into one subst_internal*)
Module subst_internal_410 (STLC: STLC_409). 
  Include tm_var_411 STLC. (* And other cases ... *)
End subst_internal_410.
Module v_33_34 (STLC: STLC_409).
(* Intermediate Module solely for type checking *)
  Include tm_rec_425 STLC.
  Parameter (recursor_for_type_checking : __recursor_type_tm_rec).
  Definition term_for_type_checking :=
    recursor_for_type_checking (fun _  ⇒ id → STLC.tm → STLC.tm)
      subst_internal_410.tm_var
      subst_internal_410.tm_abs ... .
End v_33_34.
\end{minted}
  \end{minipage}
\caption{Exempler STLC Code, especially about Inductive Type}\label{fig:plugin-example2}
\end{figure}\YZ{Change this one into STLC example, to see the layout}\EDJreply{Changed. I think it only make things worse.}

@@

However, we later still need the concrete eliminator on "tm" for doing exhaustiveness checking (as in \ref{chg:extensible-inductive-type}), thus we will store the corresponding info somewhere.  Here we show one exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module "tm_rec_425". This concrete eliminator will be used by our "FRecursor". For example, in \cref{fig:plugin-example2}, "FRecursor" command will feed Coq a module "v_33_34" for doing type-checking with specified handlers (e.g. "neg_handler") during compilation---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq. Note that, at this point of the example, "FRecursor" will only do type check. The compilation of "FRecursor" into a component of the compiled module only happens when closing the whole family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla) inductive type in that compiled module. Thus a "FRecursor" field is unlike other fields, even though it is exposed as a mundane function for the following fields in the programming interface.

@@

Importantly, the recursor field (e.g. "subst") is not actually inheritable, because it is not compiled like any other fields---recall how other fields are compiled into reusable pieces of (parametrized) modules, a recursor field is not compiled like that. For all the children family, we have to reconstruct and override the "FRecursor" field with a ``new'' recursor using (maybe the same) handler family. This is affordable since each aggregated recursor handler doesn't need
to be rechecked---look at "STLC_bool.neg_handler", there are only  new
handlers for "tm_true" and the handlers (mundane fields) from the parent
"B.subst_internal.tm_var" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "STLC_409"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "tm_var_411"---doesn't require another type-check and is re-used during the type-checking and compilation of "STLC_bool.subst". For example, for the new "subst", when compiling family "STLC_bool" into a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla but extended and modified) inductive type in that compiled module. Note that, even though "tm_rec" and "tm" has the same name as before, they are in the newly-compiled module so they refer to totally different things.

@@

\begin{minted}[escapeinside=@@]{Coq}
(* Inside Family B *)
FInductive b : Set := tt : b | ff : b.
b_prec : forall R, option R -> option R -> b -> option R.@\YZ{Not clear what it takes to define/prove partial recursors. The reader needs to be told if they require breaking abstraction like recursors do.}\EDJreply{What do you mean? recursor doesn't break abstraction but break extensibility. Below mentioned partial recursor doesn't break extensibility}@
\end{minted}
with appropriate computational axioms. This partial recursor differs from original recursor only by decorating an "option" with the return type thus possible to be automatically generated (and the case is similar for those computational axiom). The partiality doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec") because "b_prec" can just return "None" when bump into extended constructor in the future.

The first thing to check is that, if partial recursor is theoretically enough for injectiveness of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{B.tt ≠
B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : B.tt = B.ff → False.
(* Because B.tt = B.ff → reflect B.tt = reflect B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}
